---
title: "Population comparisons and incidence"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Population comparisons and incidence}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  echo = TRUE,
  warning = FALSE,
  message = FALSE,
  fig.width = 6,
  fig.height = 4
)


library(patchwork)
library(ggoutbreak)
here::i_am("vignettes/incidence-trends.Rmd")
source(here::here("vignettes/vignette-utils.R"))
```

Comparisons between epidemic states in different populations separated by for example
geography, or in this case age, requires a normalisation by the population size
in either absolute or proportional terms. Based on the 2021 census we have some
demo data for the demographics of England:

```{r}
dplyr::glimpse(england_demographics)
```

# Incidence Poisson rate model

A plot of the normalised incidence rates of COVID-19 by population size, shows
initially the rate of COVID cases was highest in the elderly. By late 2020 this
pattern had changed and rates were uniform across age groups. In early 2021 as
vaccination took hold and school testing was rolled out, younger age groups had
higher rates of COVID positive tests, with a curious spike in the very young age
groups around November 2021. In early 2022 the pattern reversed and the elderly
again became the age group with the highest rates, and that pattern has
persisted until the present. Transiently the instantaneous rates per 1000 person
years exceeded 1000 in some age groups, this is possible for limited time
periods, or if multiple episodes per year can be observed.

```{r}

tmp = ggoutbreak::england_covid %>%
   ggoutbreak::poisson_locfit_model(window=21) %>%
   ggoutbreak::normalise_incidence(ggoutbreak::england_demographics, population_unit=1000, normalise_time="1 year")

raw_pop = ggoutbreak::england_covid %>% 
  ggoutbreak::normalise_count(ggoutbreak::england_demographics, population_unit=1000, normalise_time="1 year")

plot_incidence(tmp,raw = raw_pop, size=0.25)+scale_y_log1p(n=7)+
  ggplot2::scale_colour_viridis_d(aesthetics = c("fill","colour"))
  
```

The use of test positives as a proxy for COVID incidence is clearly potentially
biased by testing (particularly in the first wave where testing was limited to those
in hospital). A more reliable comparison in this situation would be a test
positive proportion, but unfortunately testing rates are not published broken 
down by age.

The exponential growth rate is already normalised by population size. Comparisons
of the growth rate in these populations gives an idea of how tightly coupled they are.
In most age groups the epidemic is growing and shrinking in sync apart from possibly
the very young. COVID detections in this age group were not particularly reliable though
and this would be easy to over-interpret.

```{r}


plot_growth_rate(tmp)+
  ggplot2::scale_fill_viridis_d(aesthetics = c("fill","colour"))+
  ggplot2::coord_cartesian(ylim=c(-0.15,0.15))



```

The combination of growth and normalised incidence allows us to compare the
epidemic state at different time points, in this case Christmas day in 2020,
2021 and 2022. This shows the same data as the previous graphs.

```{r}

plot_growth_phase(tmp,
    timepoints = as.Date(c("Xmas 2020"="2020-12-25","Xmas 2021"="2021-12-25","Xmas 2022"="2022-12-25")),
    duration = 70, 
    interval = 7
)+
  ggplot2::scale_colour_viridis_d()

```


# Proportion model

There are two possible proportions models which would be of interest here. As
mentioned about the proportion of positive tests in each age group would give
us a clearer picture of whether the differences between age groups are down to 
differential testing, but unfortunately not available in this data set.

The second potential use is the distribution of ages in the test positive age
group. This age distribution gives us some information about where the burden of 
disease is in the population but is also biased by test prioritisation.

A multinomial proportion shows similar patterns as the normalised incidence plot
above:

```{r}


tmp2 = ggoutbreak::england_covid %>%
   ggoutbreak::proportion_locfit_model(window=21)

p1 = plot_multinomial(tmp2,normalise = TRUE)+
  ggplot2::scale_fill_viridis_d()

p2 = ggplot2::ggplot(england_demographics)+
  ggplot2::geom_bar(ggplot2::aes(x="baseline",y=population/sum(population)*100,fill=class), stat="identity", position="stack", colour="black", linewidth=0.1)+
  ggplot2::scale_fill_viridis_d(guide="none")+
  ggplot2::xlab(NULL)+
  ggplot2::ylab(NULL)+
  ggplot2::theme(axis.text.y = ggplot2::element_blank())+
  ggplot2::coord_cartesian(expand=FALSE)

p1+p2+patchwork::plot_layout(nrow=1,widths = c(20,1),guides = "collect")

```

The age distribution of test positives can be normalised by the age distribution
of the population. This give us a relative proportion of age groups in people
testing positive versus expected in the population. This is conceptually a
relative risk but of age group given COVID status 

i.e. it is $\frac{P(age = 80+|COVID+)}{P(age = 80+)}$

At any point in time for a given population this quantity is centred around 1
and comparing it to growth rate gives us a possibly clearer picture of the
trajectory of the relative distribution of COVID in the population. In Xmas 2021
although the majority of cases was in the young, relatively high growth in the
elderly population meant that they were catching up, and from above we can see
that by early 2022 the elderly had highest COVID positive rates. In 2022
however, the separation of the age groups was established and the trajectories
were acting to preserve that separation.

```{r}

tmp3 = tmp2 %>% infer_risk_ratio(england_demographics)

plot_growth_phase(tmp3,
    timepoints = as.Date(c("Xmas 2020"="2020-12-25","Xmas 2021"="2021-12-25","Xmas 2022"="2022-12-25")),
    duration = 70, 
    interval = 7
)+
  ggplot2::scale_colour_viridis_d()

```


## Pre test probability

Using the ONS COVID-19 infection survey we can look at population prevalence based on
random sampling (panel A). 

The proportion of people positive in a random sample of the population can be
compared to the proportion of people testing positive out of all the people
tested. This ratio is an indication of the pre-test probability of disease in
the non-random sample group and gives us some idea of how selectively testing is
being applied. If testing is only being done on people with symptoms for example
then their pre-test probability of disease is higher than the general
population. This is shown in panel B, and it also can be interpreted as how 
biased the sample that is tested for symptoms is.

```{r}
p1 = ggoutbreak::plot_prevalence(
    ggoutbreak::england_ons_infection_survey %>% dplyr::filter(geography == "England"),
    events = ggoutbreak::england_events
  )+
  ggplot2::theme(
    axis.title.x = ggplot2::element_blank(), 
    axis.text.x = ggplot2::element_blank(), 
    axis.text.x.bottom = ggplot2::element_blank()
  )+
  ggplot2::ylab("ONS prevalence (%)")


p2 = ggoutbreak::england_covid_pcr_positivity %>%
  proportion_locfit_model() %>%
  dplyr::inner_join(ggoutbreak::england_ons_infection_survey %>% 
                      dplyr::filter(geography=="England") %>% 
                      dplyr::rename_with(.cols = tidyselect::starts_with("prevalence"), .fn = ~stringr::str_replace(.x,"prevalence","ons")), by=c("time")
  ) %>%
  dplyr::transmute(date=date, pre_test_odds = proportion.0.5 / ons.0.5) %>%
  ggplot2::ggplot(ggplot2::aes(x=date,y=pre_test_odds)) + ggplot2::geom_line() +
  ggplot2::geom_hline(yintercept=1, colour="grey40",linetype="dashed") +
  ggplot2::ylab("Pre-test odds ratio") +
  ggoutbreak::geom_events(events = ggoutbreak::england_events,hide_labels = TRUE)+
  ggplot2::theme(
    axis.title.x = ggplot2::element_blank(), 
    axis.text.x = ggplot2::element_blank(), 
    axis.text.x.bottom = ggplot2::element_blank(), 
    axis.text.x.top = ggplot2::element_blank()
  )

```

Likewise the incidence of cases per capita can be compared to the 
prevalence. A prevalence-like estimate can be made from the incidence by convolving
the incidence estimate with a delay distribution that represents the duration of
infection. Duration of infection could be represented either by the sensitivity of a 
test to detect an infection as a function of time after infection, or as the 
probability of successful viral culture as a function of time after symptom onset.

In this case the ratio of prevalence derived from positive case counts,
resulting from symptomatic testing, and prevalence measured directly on a random
sample by the ONS survey has a connection with the ascertainment rate. The
actual value of the ascertainment rate calculated by this method depends very
heavily on assumptions on the duration of infection, and hence the prevalence
that is derived from case rates, and absolute value of the derived ascertainment
is not reliable, however the relative change in this number is still informative
and demonstrates that case ascertainment fell abruptly during the omicron wave
(panel C).


```{r fig.height=6, fig.width=8}
england_pop = ggoutbreak::england_demographics %>% dplyr::ungroup() %>% dplyr::summarise(population = sum(population))

p3 = ggoutbreak::england_covid_pcr_positivity %>% 
  ggoutbreak::poisson_locfit_model() %>%
  # N.B. `covid_viral_shedding` is technically from symptom onset not infection.
  # We could argue that this is a good thing as case positivity is closer in time
  # to symptom onset however this is forgetting about all the cases that would
  # test positive if they had been tested. An alternative to this would have been 
  # to use `covid_test_sensitivity` which looks at probability of detection from 
  # time of infection but it is only in cases with proven infection so a biased 
  # prior.
  ggoutbreak::infer_prevalence(pop = england_pop, ip = covid_viral_shedding) %>%
  dplyr::inner_join(england_ons_infection_survey %>% 
                      dplyr::filter(geography=="England") %>% 
                      dplyr::rename_with(.cols = tidyselect::starts_with("prevalence"), .fn = ~stringr::str_replace(.x,"prevalence","ons")), by=c("time")
  ) %>%
  dplyr::transmute(date=date, ascertainment = prevalence.0.5 / ons.0.5 * 100) %>%
  ggplot2::ggplot(ggplot2::aes(x=date,y=ascertainment)) + 
  ggplot2::geom_line() + 
  ggplot2::ylab("Ascertainment rate (%)") +
  ggoutbreak::geom_events(events = ggoutbreak::england_events,hide_labels = TRUE)


p1+p2+p3+patchwork::plot_layout(ncol=1)+patchwork::plot_annotation(tag_levels="A")
```



