---
title: "Infectivity profile discretisation"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Infectivity profile discretisation}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include=FALSE}
library(ggoutbreak)
library(patchwork)
utils::data("MockRotavirus",package = "EpiEstim")
si_data = MockRotavirus$si_data
si_distr = MockRotavirus$si_distr
source(here::here("vignettes/vignette-utils.R"))
```

Estimation of the infectivity profile is usually done by fitting a 
doubly censored survival model with an underlying gamma probability distribution
to data and can be done with MCMC using `EpiEstim`. In this case we demonstrate
this with a mock Rotavirus dataset from `EpiEstim`:

```{r}

dic.fit.mcmc = memoise::memoise(coarseDataTools::dic.fit.mcmc)

clever_init_param <- EpiEstim::init_mcmc_params(si_data, "G")
SI_fit_clever <- dic.fit.mcmc(dat = si_data,
  dist = "G", # Gamma distribution
  init.pars = clever_init_param,
  burnin = 1000,
  n.samples = 5000,
  verbose = 10000)

SI_fit_clever

```

The median estimate and confidence intervals are quantiles of the posterior
samples each of which define a candidate gamma distribution. The density of a
subset of these are plotted here. 

```{r}
# The sample quantiles
# apply(SI_fit_clever@samples,MARGIN = 2,FUN = quantile, p=c(0.5,0.025,0.975))

gammas = SI_fit_clever@samples %>% 
  dplyr::transmute(
    rate = 1/var2, 
    shape = var1, 
    scale = var2, 
    mean = var1*var2, 
    sd = sqrt(var1)*var2, 
    lmean = log(mean), 
    lsd = log(sd))

max_x = ceiling(stats::quantile(stats::qgamma(0.95, gammas$shape, gammas$rate),0.75))

if (is.null(si_distr)) si_distr = EpiEstim::discr_si(0:max_x, stats::quantile(gammas$mean, 0.5), stats::quantile(gammas$sd, 0.5))

ggplot2::ggplot()+
  purrr::pmap(gammas %>% utils::tail(200), 
    function(shape,rate,...) {
      ggplot2::geom_function(fun = function(x) stats::dgamma(x, shape=shape, rate=rate), alpha=0.05, xlim=c(0,max_x))
    }
  )+
  ggplot2::ylab("density")
  
```

Alternatives to generating these infectivity profile distributions from data are
examined in the vignette "Sampling the infectivity profile from published serial
interval estimates", and include resampling from published estimates.

Estimation of $R_t$ using `EpiEstim` requires that the infectivity profile
estimates are discretised. As the underlying Cori method requires that the
probability of infection at day zero is zero, `EpiEstim` discretises the
infectivity profile distributions using a offset gamma distribution which
requires that the mean of the infectivity profile distribution is greater than
1. This results in some MCMC fitted distributions being unsuitable especially in
this case where the mean is short.

If we are using a framework for estimating $R_t$ that does not have this requirement
a more natural discretisation using the the cumulative density of the
untransformed gamma distribution and cut points based on 0.5 day intervals is
possible.

```{r}

original_disc = gammas %>%
  dplyr::filter(mean > 1) %>%
  dplyr::transmute(
    coll = dplyr::row_number(),
    disc = purrr::map2(mean,sd, ~ tibble::tibble(
                         a0 = c(0,seq(0.5,length.out = 50)),
                        a = seq(0.5,length.out = 51),
                         p = EpiEstim::discr_si(0:50, .x, .y))),
    disc_type = "discr_si"
) %>% tidyr::unnest(disc)

cdf_disc = gammas %>%
    dplyr::transmute(
      discr = purrr::map2(
        shape, rate, ~ tibble::tibble(
          a0 = c(0,seq(0.5,length.out = 50)),
          a = seq(0.5,length.out = 51),
          p = dplyr::lead(stats::pgamma(a, .x ,.y),default = 1) - stats::pgamma(a, .x ,.y)
        )
      ),
      coll = dplyr::row_number(),
      disc_type = "cdf"
    ) %>% tidyr::unnest(discr)


tmp = dplyr::bind_rows(original_disc,cdf_disc) %>% 
  dplyr::mutate(disc_type=factor(disc_type, levels = c("discr_si","cdf")))

sources = length(levels(tmp$disc_type))

tmp_summ = tmp %>% dplyr::group_by(source = disc_type,a0,a) %>% dplyr::summarise(p = mean(p))

# tmp_mean = tmp %>% dplyr::group_by(source = disc_type,coll) %>% 
#   dplyr::summarise(mean = sum(a*p)) %>% 
#   dplyr::summarise(sd = stats::sd(mean),mean = mean(mean), parameter="mean") %>%
#   dplyr::bind_rows(
#     gammas %>% dplyr::summarise(sd = stats::sd(mean),mean=mean(mean), parameter="mean",source="none")
#   )

ggplot2::ggplot(tmp %>% dplyr::rename(source = disc_type))+
  ggplot2::geom_segment(mapping=ggplot2::aes(x=a0,xend=a,y=p, colour=source), alpha=0.01)+
  ggplot2::geom_segment(data = tmp_summ, mapping=ggplot2::aes(x=a0,xend=a,y=p), colour="black")+
  ggplot2::xlab("time")+
  ggplot2::coord_cartesian(xlim=c(0,max_x+1))+
  ggplot2::guides(colour = ggplot2::guide_legend(override.aes = list(alpha = 1)))+
  ggplot2::facet_wrap(~source)

# tmp_mean

```


This alternative discretisation will affect reproduction number estimation. We
might expect that for a given growth rate the reproduction number would be less
for the `cdf` discretisation as the mass of the probability looks lower. This
would mean that more generations of the epidemic can be expected in a unit time
and hence any observed exponential growth per unit time is due to fewer
secondary infections per primary infection.

We test this with a fixed value of the growth rate, and the Wallinga-Lipsitch method
of estimating reproduction number from growth rate, for both types of discretisation.
The Wallinga-Lipsitch method can use arbitrary limits for the discretisation of
the infectivity profile and hence can produce an estimate with infectivity profiles
that are impossible to use in `EpiEstim`. For comparison the result of estimating
$R_t$ using `EpiEstim`, and using the `discr_si` strategy is also included.

```{r}

R_t = tmp %>% tidyr::nest(dist = c(p,a,a0)) %>% 
  tidyr::crossing(r = c(0.4,0.2,0.1,0.05,0,-0.05,-0.1)) %>%
  dplyr::mutate(
    R_t = purrr::map2_dbl(r, dist, ~ ggoutbreak::wallinga_lipsitch(.x, y = .y$p, a0 = .y$a0, a = .y$a)),
    source = sprintf("wall-lips (%s)",disc_type)
  ) %>%
  dplyr::filter(!is.na(R_t))
  
R_t_summ = R_t %>% dplyr::group_by(r,source) %>% dplyr::summarise(
  median = stats::quantile(R_t,0.5),
  lower = stats::quantile(R_t,0.025),
  upper = stats::quantile(R_t,0.975)
)

```

```{r}
r = c(0.4,0.2,0.1,0.05,0,-0.05,-0.1)

# select a shorter list of samples
tmp2 = tmp %>% 
  dplyr::filter(disc_type == "discr_si") %>%
  dplyr::group_by(source = disc_type) %>% dplyr::reframe(
  si_matrix = list(matrix(as.vector(p),nrow=51)[,1:250]),
  r = list(r)
) %>% tidyr::unnest(r)

# tmp2$si_matrix[[1]][1:10,1:10]

cached_estimate_R = memoise::memoise(EpiEstim::estimate_R)

compare_R = tmp2 %>% dplyr::mutate(R = purrr::map2(si_matrix, r, ~ {
  ts = tibble::tibble(
        t = 0:30
      ) %>% dplyr::mutate(
        I = 100*exp(.y*t)
      )
  return(cached_estimate_R(ts, 
      method="si_from_sample", si_sample = .x,
      config = EpiEstim::make_config(t_start=2, t_end = 30))$R)
  })
)

plot2data = compare_R %>% tidyr::unnest(R) %>% dplyr::glimpse()
```

```{r}
ggplot2::ggplot(R_t_summ )+
  ggplot2::geom_point(ggplot2::aes(y=median, x = factor(r), colour = source), position=ggplot2::position_dodge(width=0.4))+
  ggplot2::geom_errorbar(ggplot2::aes(ymin=lower, ymax=upper, x = factor(r), colour = source), width=0.2,position=ggplot2::position_dodge(width=0.4))+
  ggplot2::geom_point(data = plot2data, mapping=ggplot2::aes(y=`Median(R)`, x = factor(r), colour="epiestim (discr_si)"))+
  ggplot2::geom_errorbar(data = plot2data, mapping=ggplot2::aes(ymin=`Quantile.0.025(R)`, ymax=`Quantile.0.975(R)`, x = factor(r), colour="epiestim (discr_si)"), width=0.1)+
  ggplot2::xlab("Growth rate (days⁻¹)")+ggplot2::ylab("Reproduction number")

```

We see a noticeably less extreme estimate for the reproduction number when the
infectivity profile is discretised with the unadjusted `cdf` strategy compared
with the `EpiEstim` inbuilt `discr_si` strategy. This is worse as the growth rate
increases, with `discr_si` based estimates being un to 20% higher in extreme
growth rate scenarios. The bias will be accentuated in this example as it
contains a very short serial interval, which makes the discretisation error more
apparent. We would expect to see much less bias due to discretisation if dealing
with diseases with longer serial intervals.

# Conclusion

The methods involved in `EpiEstim` require that the infectivity profile is a
discrete probability distribution and that the probability of transmission at
time zero is zero. For diseases with a shorter serial interval, these
constraints are difficult to resolve, and discretisation of the infectivity
profile distribution using a offset 1 gamma distribution results in an unusual
shape to the discrete distribution. In the Wallinga-Lipsitch framework for
estimating the reproduction number from the growth rate these constraints do not
apply and we can compare `EpiEstim`s discretisation approach to one that is
closer to the originally estimated continuous distribution, but which requires
that the framework can handle a non zero probability of transmission on day
zero. This comparison suggests that discretisation may bias `EpiEstim`s
estimates of the reproduction number by up to 20% when growth rates are high, or
even higher when compared to the equilibrium point of $R_t=1$. Alternative
strategies for discretisation and frameworks that relax the requirement for a
probability of infection of zero at time zero will benefit from further
investigation particularly for diseases with a shorter serial interval.
