---
title: "Estimating the reproduction number from modelled incidence"
output: html_document
vignette: >
  %\VignetteIndexEntry{Estimating the reproduction number from modelled incidence}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---
  
Robert Challen ^1,2^; Leon Danon ^1,2^;

1) Engineering Mathematics, University of Bristol, Bristol, UK
2) AI4CI

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  echo = FALSE,
  warning = FALSE,
  message = FALSE
)
```

```{r setup}
library(patchwork)
library(ggoutbreak)
here::i_am("vignettes/rt-from-incidence.Rmd")
source(here::here("vignettes/vignette-utils.R"))
```


# Introduction

If we have estimated the incidence of a disease $I_t$ using a poisson rate using
maximum likelihood estimators, the rate is typically a log-normally distributed
with parameters $\mu$ and $\sigma$. Such a fitted model is shown below on a log1p
scale, for the COVID-19 epidemic in England:

```{r}

raw = ggoutbreak::england_covid %>%
  time_aggregate(count=sum(count))

fit = raw %>% poisson_locfit_model()

plot_incidence(fit,raw, colour="blue",size=0.025,events = england_events)+
  scale_y_log1p()+
  ggplot2::coord_cartesian(xlim=as.Date(c("2020-01-01","2023-01-01")))
```

It is appealing to use this modelled incidence estimate to calculate an estimate
of the reproduction number, $R_t$. Incidence models can be derived in a number
of ways, they are easily inspected for error and can be made tolerant of missing
values and outliers. 

# Methods

To use a modelled estimate of incidence to predict $R_t$ we
need to propagate uncertainty in incidence into our $R_t$ estimates. To
calculate $R_t$ we can use the backwards-looking renewal equations which
incorporate the infectivity profile of the disease ($\omega$) at a number of
days after infection ($\tau$):


$$
I_t \sim Lognormal(\mu_t,\sigma_t) \\
R_t = \frac{I_t}{\sum_{\tau}{\omega_{\tau}I_{t-\tau}}}
$$

giving us:

$$
R_t = \frac{Lognormal(\mu_t,\sigma_t)}{\sum_{\tau}{
  Lognormal( \mu_{t-\tau} + log(\omega_{\tau}) , \sigma_{t-\tau})
}} \\
$$


The sum of $i$ such log normal distributions can be approximated by another log
normal (Lo 2013) with parameters $\mu_Z$ and $\sigma_Z$.

$$
\begin{align}
	S_+ &= \operatorname{E}\left[\sum_i X_i \right] = \sum_i 
	\operatorname{E}[X_i] = 
	\sum_i e^{\mu_i + \frac{1}{2}\sigma_i^2}
	\\
	\sigma^2_{Z} &= \frac{1}{S_+^2} \, \sum_{i,j}
	  \operatorname{cor}_{ij} \sigma_i \sigma_j \operatorname{E}[X_i] \operatorname{E}[X_j] =
	  \frac{1}{S_+^2} \, \sum_{i,j}
	  \operatorname{cor}_{ij} \sigma_i \sigma_j e^{\mu_i+\frac{1}{2}\sigma_i^2} 
	  e^{\mu_j+\frac{1}{2}\sigma_j^2} 
	\\
	\mu_Z &= \ln\left( S_+ \right) - \frac{1}{2}\sigma_{Z}^2 
\end{align}
$$

The sum term in the denominator of the renewal equations consists of a set of
correlated scaled log normal distributions with both scale and correlation
defined by the infectivity profile ($\omega$). In our case $cor_{ij}$ can be
equated to the infectivity profile ($\omega_{|i-j|}$) when $i \neq j$ and 1 when
$i = j$. $\mu_i$ is $\mu_{t-\tau} + ln(\omega_{\tau})$.


$$
\begin{align}
	S_{t} &= \sum_{s=1}^{|\omega|} { \omega_s e^{\mu_{t-s} + \frac{1}{2}\sigma_{t-s}^2 }} \\
	\sigma_{Z,t} &= \sqrt{
	  \frac{
	    \sum_{i,j=1}^{|\omega|} {
  	    (\omega_{|i-j|}+I(i,j)) \omega_i \omega_j (\sigma_{(t-i)} e^{\mu_{(t-i)}+\frac{1}{2}\sigma_{(t-i)}^2}) (\sigma_{(t-j)} e^{\mu_{(t-j)}+\frac{1}{2}\sigma_{(t-j)}^2}) 
	    }
	  }{S_{t}^2}
	}	\\
	\mu_{Z,t} &= \log\left( S_{t} \right) - \frac{1}{2}\sigma_{Z,t}^2 
\end{align}
$$

$\mu$ is the central estimate of case counts on the log scale, and its standard
deviation can also be large. There are numerical stability issues dealing
with terms involving $e^{(\mu+\sigma^2)}$, however keeping everything in log space and 
using optimised log-sum-exp functions this can be made computationally tractable.

$$
\begin{align}
	\log(S_{t}) &= \log(\sum_{s=1}^{|\omega|} {  e^{\mu_{t-s} + \frac{1}{2}\sigma_{t-s}^2 + \log(\omega_s) }}) \\
	\log(T_{t,\tau}) &= \log(\omega_{\tau}) + \log(\sigma_{(t-{\tau})}) + \mu_{(t-{\tau})} + \frac{1}{2}\sigma_{(t-{\tau})}^2) \\
	\log(cor_{i,j}) &= \log(\omega_{|i-j|}+I(i=j)) \\
	\log(\sigma_{Z,t}^2) &= \log(
	    \sum_{i,j=1}^{|\omega|} {
  	      e^{
  	        \log(cor_{i,j}) + \log(T_{t,i}) + \log(T_{t,j})
  	      }
	    }) - 2 \log(S_{t}) \\
	\mu_{Z,t} &= \log( S_{t} ) - \frac{1}{2}\sigma_{Z,t}^2 
\end{align}
$$

N.B. if we assume the individual estimates of the incidence are uncorrelated this
simplifies to:

$$
\begin{align}
\log(\sigma_{Z,t}^2) &= \log(
	    \sum_{\tau=1}^{|\omega|} {
  	      e^{
  	        2 \log(T_{t,\tau})
  	      }
	    }) - 2 \log(S_{t})
\end{align}
$$

Empirically there is not a huge amount of difference in estimates between these
two forms. If the infectivity profile $\omega$ is spread out over a large period
then the correlation matrix will be $O(\omega)^2$ which may predicate this simpler
order 1 formulation.

With $\mu_{Z,t}$ and $\sigma_{Z,t}$ we are left with the final derivation of
$R_t$, giving us a distributional form of $R_t$ incorporating uncertainty from
modelled incidence estimates:

$$
\begin{align}
R_t &= \frac{Lognormal(\mu_t,\sigma_t)}
{Lognormal( \mu_{Z,t}, \sigma_{Z,t})} \\
\mu_{R_t} &= \mu_t - \mu_{Z,t} \\
\sigma_{R_t} &= \sqrt{\sigma_t^2+\sigma_{z,t}^2} \\
R_t &= Lognormal(\mu_{R_t}, \sigma_{R_t})
\end{align}
$$

This is conditioned on a single known infectivity profile ($\omega$). In reality there is
also uncertainty in the infectivity profile, however we cannot assume any
particular distributional form for this. We can use a set of empirical
estimates of the infectivity profile ($\Omega$) to calculate multiple distributional
estimates for the reproduction number ($R_{t,\omega}$) and then combine these as a mixture distribution.

There is not much we can say about this mixture distribution, as it will depend on the 
nature of the various empirical infectivity profile distributions. However, we can 
use general properties of mixture distributions to generate estimates for the 
mean and variance of the reproduction number including uncertainty 
arising from multiple infection profile estimates ($R_t^*$):

$$
\begin{align}
E[R_t|\omega] &= e^{(\mu_{R_t,\omega} - \frac{1}{2}\sigma_{R_t,\omega}^2)} \\
V[R_t|\omega] &= \big[
  e^{(\sigma_{R_t,\omega}^2)} - 1
\big] \big[
  e^{2 \mu_{R_t,\omega} + \sigma_{R_t,\omega}^2}
\big] \\
E[R_t^*] &= \frac{1}{|\Omega|}\sum_{\omega \in \Omega} E[{R_t|\omega}] \\
V[R_t^*] &= \frac{1}{|\Omega|} \bigg(\sum_{\omega \in \Omega}{V[R_t|\omega]+E[R_t|\omega]^2}\bigg) - E[R_t^*]^2 \\
\end{align}
$$

The cumulative distribution function of the mixture is simply the arithmetic mean of the 
component cumulative distribution functions (conditioned on each infectivity profile). If
$\Phi$ is the cumulative distribution function of the standard normal distribution:


$$
\begin{align}
F_{R_t^*}(x) &= \frac{1}{|\Omega|}\sum_{\omega \in \Omega}F_{R_t}(x|\omega) \\
P(R_t^* \le x) &= \frac{1}{|\Omega|}\sum_{\omega \in \Omega} P(R_{t,\omega} \le x) \\
P(R_t^* \le x) &= \frac{1}{|\Omega|}\sum_{\omega \in \Omega} \Phi\bigg(\frac{ln(x) - \mu_{R_t,\omega}}{\sigma_{R_t,\omega}}\bigg)
\end{align}
$$

As the cumulative density function of this mixture distribution is a strictly increasing function
specific solutions for median and 95% confidence intervals can be calculated
numerically by solving the following equations:

$$
\begin{align}
\frac{1}{|\Omega|}\sum_{\omega \in \Omega} \Phi\bigg(\frac{ln(q_{0.025}) - \mu_{R_t,\omega}}{\sigma_{R_t,\omega}}\bigg) - 0.025 &= 0 \\
\frac{1}{|\Omega|}\sum_{\omega \in \Omega} \Phi\bigg(\frac{ln(q_{0.5}) - \mu_{R_t,\omega}}{\sigma_{R_t,\omega}}\bigg) - 0.5 &= 0 \\
\frac{1}{|\Omega|}\sum_{\omega \in \Omega} \Phi\bigg(\frac{ln(q_{0.975}) - \mu_{R_t,\omega}}{\sigma_{R_t,\omega}}\bigg) - 0.975 &= 0
\end{align}
$$

# Implementation

This method is implemented using the following R function, which is designed
for numerical stability and speed. Generating $R_t$ estimates given modelled 
incidence typically occurring in:

```{r}
print(ggoutbreak:::.internal_r_t_estim)
```

# Results

Testing this against the incidence model shown above, and comparing the results
to the SPI-M-O consensus $R_t$ estimates gives us the following time-series for
England. This has not been formally evaluated but qualitatively is a good fit. This 
single time series with 1410 time points took a few seconds to calculate the 
reproduction number from , which opens
up the possibility of performing $R_t$ estimates in fine grained geographical 
or demographic subgroups.

```{r}

system.time({
  rt_fit = fit %>% ggoutbreak::rt_from_incidence(ip = covid_ip)
})

plot_rt(rt_fit, events = england_events, colour="blue")+
  ggplot2::coord_cartesian(xlim=as.Date(c("2020-01-01","2023-01-01")), ylim=c(0.6,1.6))+
  ggplot2::geom_errorbar(data=england_consensus_rt,ggplot2::aes(x=date-14,ymin=low,ymax=high),colour="red")

```

# Conclusion

We present a methodology for deriving $R_t$ from modelled estimates of incidence
while propagating uncertainty. We demonstrate it produces satisfactory
qualitative results against COVID-19 data. This method is relatively quick,
fully deterministic, and can be used on top of any statistical models for
estimating incidence which use logarithmic link functions.
