---
title: "Data wrangling and working with `ggoutbreak`"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Data wrangling and working with `ggoutbreak`}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}

knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  echo = TRUE,
  warning = FALSE,
  message = FALSE,
  fig.width = 6,
  fig.height = 4
)

library(ggoutbreak)
here::i_am("vignettes/time-periods.Rmd")
source(here::here("vignettes/vignette-utils.R"))

```


`ggoutbreak` assumes a consistent naming scheme for significant columns,
notably `time` and `count`, and additionally `class`, `denom`, `population`
columns. Data needs to be supplied using these column names and to make sure the
data is formatted correctly it undergoes quite rigorous checks. One area where
this can pose problems is in correct grouping, which must make sure that each
group is a single time series of unique `time` and minimally `count` columns.

# Line lists vs. time series

Infectious disease data usually either comes as a set of observations of an
individual infection with a time stamp (i.e. a line list) or as a count of
events (e.g. positive tests, hospitalisations, deaths) happening within a
specific period (day, week, month etc.) as a time series.

For count data there may also be a denominator known. For testing this could be
the number of tests performed, or the number of patients at risk of
hospitalisation.

For both these data types there may also be a class associated with each
observation, defining a subgroup of infections of interest. This could be the
variant of a virus, or the age group, for example. It may make sense to compare
these different subgroups against each other. In this case the denominator may
be the total of counts among all groups per unit time. Additionally there may be
information about the size of the population for each subgroup.

`ggoutbreak` assumes for the most part that the input data is in the form of
a set of time series of counts, each of which has a unique set of times, which
are usually complete. To create datasets like this from line lists `ggoutbreak`
provides some infrastructure for dealing with time series

# Coercing data into `ggoutbreak` format

## A simple linelist

At the lowest level a linelist might be the list of timestamps of some 
observation. This is the easiest input to process and is imported into a single
column data frame. 

The linelist is timestamped using a S3 `time_period`
class which is simply a numeric of defined length time units (in this case days)
from an origin (in this case `2019-12-29`). We can plot the individual cases.

```{r}

# Set the default start date and unit for `ggoutbreak`:
# This is not absolutely required as will set itself but helps in the long run
set_defaults("2019-12-29","1 day")

# Timestamps of 100 cases in first 40 days with a exponential growth rate of 0.1.
lldates = as.Date("2019-12-29")+rexpgrowth(100,0.1,40)
ll = lldates %>% linelist() 
ll %>% dplyr::glimpse()
ll %>% plot_cases(individual=TRUE)

```

Supposing we had a slightly more complex input data. Maybe we also had gender 
and age as well as a `onset_dates` column in a dataframe. We can let `ggoutbreak`
guess the structure of the linelist, and time stamp it. If there are multiple date columns we could also have specified the date:

```{r}

lldf = dplyr::tibble(
  onset_dates = lldates,
  gender = rcategorical(100, c(male = 0.54, women = 0.46), factor = TRUE),
  age = rcategorical(100, c(`0-17` = 0.2, `18-64` = 0.6, `65+`=0.2), factor = TRUE)
)

# in this case we can let `ggoutbreak` guess the structure of this linelist:

ll2 = lldf %>% linelist()

ll2 %>% dplyr::glimpse()

# And we can plot it with the data ordered how we like:
ll2 %>% 
  dplyr::arrange(dplyr::desc(age)) %>%
  plot_cases(mapping = ggplot2::aes(fill=age), individual=TRUE)

ll2 %>% 
  dplyr::arrange(gender) %>%
  plot_cases(mapping = ggplot2::aes(fill=gender), individual=TRUE)

```

```{r}

# In this case as it is a different time we can if we like anchor the time
# series to the `start` so the time column runs from zero. In this case we are
# also going to make the time stamp count in weeks:

sl2014 = outbreaks::ebola_sierraleone_2014 %>% 
  linelist(date = date_of_onset, anchor="start", unit="weeks") %>%
  dplyr::mutate(
    age_cat = cut(age, c(0,18,65,Inf)),
    dplyr::across(dplyr::where(is.character),forcats::as_factor)
  ) %>%
  dplyr::glimpse()

```

Plotting this is also easy, and in this case it is organised by week:

```{r}
sl2014 %>%
  dplyr::arrange(status) %>%
  plot_cases(mapping = ggplot2::aes(fill=status))
```


## Count data

Count data describes the number of cases in a time period, e.g. a day, a week 
etc. It may also be associated one or more major grouping, such as age category, gender, geographical region, ethnicity, viral subtype. In all these groups there 
is potentially a population denominator, but often there is a major grouping 
that may define a different denominator (e.g. total tested individuals). These categories can be arbitrarily bought into `ggoutbreak` and managed as groups in 
exactly the same was as you would in a normal `dplyr` pipeline.

Because count data needs to be unique for each group for each time point extra
checks are carried out to make sure that the count grouping is correct:

```{r}

nhs111 = outbreaks::covid19_england_nhscalls_2020 %>%
  dplyr::group_by(sex,ccg_code,ccg_name,nhs_region) %>%
  timeseries(
    class = age
  ) %>%
  dplyr::glimpse()

```

In this case the level of detail may be more than we want to deal with and 
aggregating groups maybe useful (mre details later):

```{r}
nhs111combined = nhs111 %>% 
  dplyr::group_by(sex,class,age) %>%
  time_aggregate() %>% 
  dplyr::glimpse()

nhs111combined %>% plot_counts(mapping=ggplot2::aes(colour=class))+
  ggplot2::facet_wrap(~sex)+
  scale_y_log1p()
```

# More detail on time periods

A weekly case rate represents a time slice of seven days with a start and finish
date. Dates are a continuous quantity, and `cut_dates()` can be used to classify
continuous dates into periods of equal duration, with a start date:

```{r}
random_dates = Sys.Date()+sample.int(21,50,replace = TRUE)
cut_date( random_dates, unit = "1 week", anchor = "start", dfmt = "%d %b")
```

Performing calculations using interval censored dates is awkward. A numeric
version of dates is useful that can keep track of both the start date of a time
series and its intrinsic duration, as a numeric. This is the purpose of the
`time_period` class:

```{r}
dates = seq(as.Date("2020-01-01"),by=7,length.out = 5)
tmp = as.time_period(dates)
tmp
```

The `time_period` defaults to using the start of the first defined `time_period`
in a session as its origin and calculating a duration unit based on the data (in
this case weekly). This behaviour can be controlled explicitly with
`set_default_start()` and `set_default_unit()`, or on an ad-hoc basis by
providing the `start_date` or `anchor` and `unit` variables to calls to
`as.time_period()`

A usual set of S3 methods are available such as formatting, printing, labelling,
and casting `time_period`s to and from dates and `POSIXct` classes:

```{r}
suppressWarnings(labels(tmp))
```

A weekly time series can be recast to a different frequency, or start date:

```{r}

tmp2 = as.time_period(tmp, unit = "2 days", start_date = "2020-01-01")
tmp2

```

and the original dates should be recoverable:

```{r}
as.Date(tmp2)
```

`date_seq()` can be used to make sure a set of periodic times is complete:

```{r}

tmp3 = as.time_period(Sys.Date()+c(0:2,4:5)*7,anchor = "start")
as.Date(date_seq(tmp3))

```

`time_period`s can also be used with monthly or yearly data but such data are
not regular. This is approximately handled and irregular date periods are
generally OK to use with `ggoutbreak`. Some functions like `date_seq` may not
work as anticipated with irregular dates, and some conversions between weeks and
months, for example, are potentially lossy.

Two time series can be aligned to make them comparable, although by default the
creation of time series means they are already comparable:

```{r}

orig_dates = Sys.Date()+1:10*7

# a 2 daily time series based on weekly dates
t1 = as.time_period(orig_dates, unit = "2 days", anchor = "2021-01-01")
t1

# a weekly with different start date
t2 = as.time_period(orig_dates, unit = "1 week", anchor = "2022-01-01")
t2

# rebase t1 into the same format as t2
# as t1 and t2 based on the same original dates converting t2 onto the same
# peridicty as t1 results in an identical set of times
t3 = as.time_period(t1,t2)
t3

# This happens automatically when the vectors are concatented
c(t1,t2)

```

# Times in `ggoutbreak` and conversion of line-lists

`ggoutbreak` uses the `time_period` class internally extensively. Casting dates
to and from `time_periods` is all that generally needs to be done before using 
`ggoutbreak` and this is done automatically by `linelist()` and `timeseries()` functions. Most of the functions in `ggoutbreak` operate on time series data
which expect a unique (and usually complete) set of data on a periodic time.

To help prepare line-list data into time series there is the `time_summarise()` 
function. A minimal line-list will have a date column and nothing else. From 
this we can generate a count of cases over a specific time unit:


```{r}

random_dates = Sys.Date()+sample.int(21,50,replace = TRUE)
linelist = dplyr::tibble(date = random_dates)
linelist %>% time_summarise(unit="1 week") %>% dplyr::glimpse()

```

If the line-list contains a `class` column it is interpreted as a complete
record of all possible options from which we can calculate a denominator. In
this case the positive and negative results of a test:

```{r}

random_dates = Sys.Date()+sample.int(21,200,replace = TRUE)
linelist2 = dplyr::tibble(
  date = random_dates,
  class = stats::rbinom(200, 1, 0.04) %>% ifelse("positive","negative")
)
linelist2 %>% time_summarise(unit="1 week") %>% dplyr::glimpse()


```

In this specific example subsequent analysis with `ggoutbreak` may focus on the
`positive` subgroup only, as the comparison between `positive` and `negative`
test results is trivial. In another example `class` may not be test results, it
could be any other major subdivision e.g. the variant of a disease. In this case
the comparison between different groups may be much more relevant. The use of
`class` as the major sub-group is for convenience. Additional grouping other
than `class` columns is also possible for multi-faceted comparisons, and
grouping is preserved but not included automatically in the denominator, which
may need to be manually calculated:

```{r}

random_dates = Sys.Date()+sample.int(21,200,replace = TRUE)
variant = apply(stats::rmultinom(200, 1, c(0.1,0.3,0.6)), MARGIN = 2, function(x) which(x==1))

linelist3 = dplyr::tibble(
  date = random_dates,
  class = c("variant1","variant2","variant3")[variant],
  gender = ifelse(stats::rbinom(200,1,0.5),"male","female")
)
  
count_by_gender = linelist3 %>% 
  dplyr::group_by(gender) %>% 
  time_summarise(unit="1 week") %>% 
  dplyr::arrange(time, gender, class) %>%
  dplyr::glimpse()

```

# Aggregating time series datasets.

In the case of a time series with additional grouping present, removing a level
of grouping whilst retaining time is made easier with `time_aggregate()`. In
this case we wish to sum `count` and `denom` by gender, retaining the class
grouping.

```{r}

count_by_gender %>% 
  dplyr::group_by(class,gender) %>% 
  time_aggregate() %>%
  dplyr::glimpse()

```

by default `time_aggregate` will sum any of `count`, `denom` and `population`
columns but any other behaviour can be specified by passing `dplyr::summarise`
style directives to the function.


