---
title: "Estimating the reproduction number from weekly data"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Estimating the reproduction number from weekly data}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

# Background

```{r setup, include=FALSE}

knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  echo = TRUE,
  warning = FALSE,
  message = FALSE,
  fig.width = 6,
  fig.height = 4
)

library(patchwork)
library(ggoutbreak)
here::i_am("vignettes/weekly-incidence.Rmd")
source(here::here("vignettes/vignette-utils.R"))
```

It is common for epidemic data to be aggregated into weekly, monthly or even
yearly case counts, for example the result of batching in reporting processes.
`ggoutbreak` has a set of infrastructure to allow case counts to be used to
estimate the reproduction number, regardless of their periodicity.

# Simulation with weekly data

We construct a very simple count model based on a time-varying growth rate which
is specified as an exponential growth in case counts per week. The `time_unit`
is defined here as 1 week, but this would typically be determined by the data.
`ggoutbreak` will generally make an informed guess at the periodicity of data
when we construct a `time_period` column from dates, if it is not stated.

In this case the simulation generates an outbreak over the course of one year
with a maximum number of weekly cases around 250, resulting from an initial
import of 30 infectious cases in week 0.

```{r}

changes = tibble::tribble(
  ~t,    ~r,
  0,      0,
  5,     0.2,
  10,     0.1,
  20,     0,
  25,     -0.1,
  35,     -0.05,
  45,     0.1
)
  
sim = sim_poisson_model(
  changes = changes,
  max_time = 52,
  time_unit = "1 week",
  fn_imports = ~ ifelse(.x == 0, 30, 0),
  seed=100
)

events = attr(sim,"events")

plot_cases(sim, events = events, colour="grey60")

```

We can go ahead and model the case rate and exponential growth rate using a
poisson model and plot them, with the parametrised growth rate in red. In this
case the time unit for both rates is per week. Although the time unit is per
week, the model output is creating an estimate per day (controlled by the
`frequency` parameter here), to allow the smooth plot.

```{r}

# Estimate weekly incidence and weekly exponential growth rate
incidence = sim %>% ggoutbreak::poisson_locfit_model(frequency = "1 day",deg = 2, window=5)

p1 = plot_incidence(incidence, events=events, raw = sim)
p2 = plot_growth_rate(incidence,events = events)+ggplot2::coord_cartesian(ylim=c(-0.05*7,+0.05*7))+
  # The simulation data includes the parametrised growth rate:
  ggplot2::geom_line(data = sim, mapping = ggplot2::aes(x=as.Date(time), y=growth), colour="red")
p1+p2+patchwork::plot_layout(ncol=1,axes="collect")+patchwork::plot_annotation(tag_levels = "A")

```

We want to estimate the reproduction number, and for this our methods need
estimates of the daily incidence of cases, as input to the renewal equation. To 
do this we rescale the temporal dimension of the model (using `rescale_model()`) 
and we replot it. In this plot the daily case rate is estimated to peak at just 
over 30 (= 250/7 approx). The growth rate is also rescaled to account for the
fact that the exponential growth rate per day is compounded up in the per week
figures above.

```{r}

# Rescale weekly to daily incidence estimates
incidence2 = incidence %>% rescale_model(time_unit = "1 day")

p1 = plot_incidence(incidence2, events=events, raw = sim)
p2 = plot_growth_rate(incidence2,events = events)+ggplot2::coord_cartesian(ylim=c(-0.05,+0.05))

p1+p2+patchwork::plot_layout(ncol=1,axes = "collect")+patchwork::plot_annotation(tag_level="A")
```

From this we can easily use our incidence estimates and an generation time 
distribution to estimate the reproduction number for this time-series. To compare
it we use a different method (Wallinga-Lipsitch 2007) to convert the growth rate
parametrisation direct to an equivalent reproduction number. We see good 
agreement, and the uncertainty associated with $R_t$ estimates derived from
weekly count data is appropriately large, and include uncertainty in the 
generation time.

```{r}

# Convert weekly growth rate input to simulation to a reproduction number
# and ensure time scales align.
changes2 = changes %>% dplyr::transmute(
  t = t*7,
  R = wallinga_lipsitch(r/7, ganyani_ip_2)
)
rt_for_date = \(t) cfg_step_fn(changes2)(as.time_period(t, unit="1 day"))

# Calculate the reproduction number from daily incidence estimates
rt_estim = incidence2 %>% rt_from_incidence(ip = ganyani_ip_2,approx = TRUE)

plot_rt(rt_estim,events = events)+
  ggplot2::geom_function(fun=rt_for_date,colour="red",n=nrow(rt_estim))

```
